---
title: "Pick out Quantitative traits"
output: html_document
date: "2025-09-25"
---

https://cran.r-project.org/web/packages/zscorer/readme/README.html

###############################################################################################
# library
###############################################################################################

```{r packages}

library(bigrquery)
library(tidyverse)
library(dplyr)
library(devtools)
library(data.table)
library(ggplot2)
library(cowplot)
library(zscorer)
library(patchwork)
library(ggrepel)


```

###############################################################################################
# read in data
###############################################################################################

https://docs.google.com/document/d/10W2jN5oHvuWXFANV4zi8RcChNYmE7ZzIsb47PPBex3k/edit?tab=t.65b2fqbcqh8a

wget https://raw.githubusercontent.com/manning-lab/primed_consortia_legacy_project_phenotypes/main/ICD-10-Pregnancy_AB_22NOV2024.csv

wget https://github.com/manning-lab/primed_consortia_legacy_project_phenotypes/raw/main/ICD-10CM-Pregnancy_Updated_AB_25Oct2024.csv

wget https://github.com/manning-lab/primed_consortia_legacy_project_phenotypes/raw/main/ICD-9CM-Pregnancy_Updated_AB_25Oct2024.csv

wget https://github.com/manning-lab/primed_consortia_legacy_project_phenotypes/raw/main/ICD-9CM-Cancer_AB_23Oct2024.csv

wget https://github.com/manning-lab/primed_consortia_legacy_project_phenotypes/raw/main/ICD-10CM-Cancer_AB_23Oct2024.csv

wget https://github.com/manning-lab/primed_consortia_legacy_project_phenotypes/raw/main/ICD-10-All_Cancer_AB_22NOV2024.csv

```{r read phenotypes}

phecodes_binary <- read.csv( "../20250908_phenotypes_v2/phecodes_binary_by_indv.csv")

```

```{r cancer and pregnancy icd codes}


icd10_preg_nov <- read.csv("ICD-10-Pregnancy_AB_22NOV2024.csv")

icd10_preg_oct <- read.csv("ICD-10CM-Pregnancy_Updated_AB_25Oct2024.csv")



icd9_preg_oct <- read.csv("ICD-9CM-Pregnancy_Updated_AB_25Oct2024.csv")



icd9_canc_oct <- read.csv("ICD-9CM-Cancer_AB_23Oct2024.csv")



icd10_canc_oct <- read.csv("ICD-10CM-Cancer_AB_23Oct2024.csv")

icd10_canc_nov <-read.csv("ICD-10-All_Cancer_AB_22NOV2024.csv")


```

```{r}

phecodes <- read.csv("../20250908_phenotypes_v2/total_enounter_icd_phe.csv")

```

```{r icd to phecode map}

phecode_map <- read.csv("../20250801_arcus_phenotype_counts/Phecode_map_v1_2_icd9_icd10cm.csv")

```

```{r get project id}

bq_auth()
project_id <- bq_projects()[1]

```

```{r encounter}

query_encounter <- "
SELECT
  pat_id,
  encounter_id,
  effective_age,
  admit_source_code,
  admit_source_name,
  height_raw,
  height_cm,
  weight_kg,
  weight_oz,
  head_circumference
FROM
  arcus.encounter
"
encounter_df <- bq_project_query(project_id, query_encounter) %>%
  bq_table_download(bigint = "integer64") %>%
  as_tibble()
head(encounter_df)

```

 c("pat_id", "encounter_id", "effective_age","admit_source_code", "admit_source_name", "height_raw", "height_cm", "weight_kg", "weight_oz", "head_circumference", )
 
```{r}

query_patient <- "
SELECT
  pat_id,
  sex,
  race,
  ethnicity,
  birth_length_in,
  birth_length_cm,
  birth_weight_oz,
  birth_weight_kg,
  gestational_age_num,
  dod_age,
  country,
  state
FROM
  arcus.patient

"
patient_df <- bq_project_query(project_id, query_patient) %>%
  bq_table_download(bigint = "integer64") %>%
  as_tibble()

head(patient_df)

```

pat_id,
sex,
race,
ethnicity,
birth_length_in,
birth_length_cm,
birth_weight_oz,
birth_weight_kg,
gestational_age_num,
dod_age,
country,
state

###############################################################################################
# encounter
###############################################################################################

```{r}

# list of columns to check
cols_to_check <- c("admit_source_code", "admit_source_name", "height_raw", "height_cm", "weight_kg", "weight_oz", "head_circumference")

# keep only rows where at least one is not NA
encounter_df <- encounter_df[rowSums(!is.na(encounter_df[cols_to_check])) > 0, ]


```

```{r}

length(unique(encounter_df[!(is.na(encounter_df$head_circumference)),]$pat_id))

```

```{r}

encounter_df <- left_join(encounter_df, patient_df[,c("pat_id", "sex")], by = "pat_id")

```

```{r}

encounter_df$sex_num <- ifelse(encounter_df$sex == "Male", 1,
                               ifelse(encounter_df$sex == "Female", 2, NA))

```

```{r}

encounter_df$age_months <- floor(encounter_df$effective_age / 30.4375)  

```

Body mass index-for-age (bfa) z-scores for children aged between zero and 228 months
age needs to be in days for this
```{r calculate bmi-age-z}

encounter_df <- addWGSR(data = encounter_df, sex = "sex_num", firstPart = "weight_kg", 
               secondPart = "height_cm", thirdPart = "effective_age", index = "bfa", 
               output = "bmiAgeZ", digits = 4)  
```

Head circumference-for-age (hca) z-scores for children aged between zero and 60 months
head_circumference in cm
```{r}

encounter_df <- addWGSR(data = encounter_df, sex = "sex_num", firstPart = "head_circumference", 
               secondPart = "effective_age", index = "hca", 
               output = "hcAgeZ", digits = 4)  
```


BMI from sandra: 

To consider an individual’s body mass index measurement, we first implemented the following measurement exclusion criteria:
Excluded measurements for which age at measurement was less than 18 years
Excluded measurement taken at inpatient or emergency department visits. Alternatively, kept only measurements taken at outpatient visits.
Excluded measurements that had cancer or pregnancy codes at the time of BMI measurement
Cancer
Pregnancy
After applying the above exclusions, we removed measurements that were outside +/- 6SD from the overall mean BMI in the total population.
Once all above exclusions were applied, we used the remaining measurements to compute the final value for each individual by taking the median BMI. Age was calculated using the median age at measurement from the remaining measurements.

https://phekb.org/phenotype/body-mass-index-bmi-0

Weight for Age and Height for Age are legacy functions and use whole months for their calculations

Weight-for-age (wfa)
```{r wfa}

encounter_df <- addWGSR(data = encounter_df, sex = "sex_num", firstPart = "weight_kg", 
               secondPart = "effective_age", index = "wfa")
```

```{r remove NA's fr hfa}

hfa_df <- na.omit(encounter_df[,c("pat_id", "encounter_id", "effective_age", "sex_num", "height_cm")])

```

```{r filter by age}

hfa_df <- hfa_df[hfa_df$effective_age >= (24*30.4375) & hfa_df$effective_age <= (228*30.4375), ]

```

Height-for-age (hfa)
```{r hfa}

encounter_df <- addWGSR(data = encounter_df, sex = "sex_num", firstPart = "height_cm", 
               secondPart = "age_months", index = "hfa")

```

```{r hfa}

hfa_df <- addWGSR(data = hfa_df, sex = "sex_num", firstPart = "height_cm", 
               secondPart = "effective_age", index = "hfa")

```

```{r}

getWGS(sexObserved = 1,
              firstPart = 86.49970,      # Height in centimetres
              secondPart = 40,
              index = "hfa")   

```


Weight-for-length (wfl)
```{r wfl}


```

Weight-for-height (wfh)
```{r wfh}


```

```{r}

write_csv(encounter_df, "encounter_quant.csv")

```

```{r}

hist(encounter_df$bmiAgeZ,
     breaks = 30,         # more bins to see detail
     main = "Distribution of BMI-for-Age Z-scores",
     xlab = "BMI z-score",
     col = "skyblue")

```

###############################################################################################
# Patient DF
###############################################################################################

```{r}

patient_df$birth_age <- 0

```

```{r}

patient_df$sex_num <- ifelse(patient_df$sex == "Male", 1,
                               ifelse(patient_df$sex == "Female", 2, NA))

```

```{r}

patient_df <- addWGSR(data = patient_df, sex = "sex_num", firstPart = "birth_weight_kg", 
               secondPart = "birth_length_cm", thirdPart = "birth_age", index = "bfa", 
               output = "bmiAgeZ", digits = 4)  
```

```{r}

patient_df$gestational_age_num <- as.numeric(patient_df$gestational_age_num)

```

```{r}

# function to create a pie chart for any column
make_pie <- function(data, column_name) {
  data %>%
    filter(!is.na(.data[[column_name]])) %>%
    count(!!sym(column_name)) %>%
    ggplot(aes(x = "", y = n, fill = !!sym(column_name))) +
    geom_col(width = 1) +
    coord_polar(theta = "y") +
    theme_void() +
    labs(title = paste("Distribution of", column_name), fill = column_name)
}

```

```{r}

make_pie <- function(data, column_name) {
  df <- data %>%
    filter(!is.na(.data[[column_name]])) %>%
    count(!!sym(column_name)) %>%
    arrange(desc(n)) %>%
    mutate(
      prop = n / sum(n),
      cumulative = cumsum(prop),
      midpoint = cumulative - prop / 2,
      angle = 2 * pi * midpoint,
      # radius for text outside the pie
      label_x = 1.2 * sin(angle),
      label_y = 1.2 * cos(angle)
    )

  ggplot(df, aes(x = "", y = prop, fill = !!sym(column_name))) +
    geom_col(width = 1, color = "white") +
    coord_polar(theta = "y") +
    geom_text(
      aes(x = label_x, y = label_y, label = n),
      color = "black", size = 3.5, fontface = "bold"
    ) +
    theme_void() +
    labs(
      title = paste("Distribution of", column_name),
      fill = column_name
    )
}


```

```{r}

# make a pie chart for each column
pie_race <- make_pie(patient_df, "race")
pie_ethnicity <- make_pie(patient_df, "ethnicity")
pie_country <- make_pie(patient_df, "country")
pie_state <- make_pie(patient_df, "state")
pie_sex <- make_pie(patient_df, "sex")

```

```{r}

ggsave("acrus_race.pdf", plot = pie_race, width = 6, height = 4, units = "in")

```

```{r}

ggsave("acrus_sex.pdf", plot = pie_sex, width = 6, height = 4, units = "in")

```

```{r}

ggsave("acrus_state.pdf", plot = pie_state, width = 6, height = 4, units = "in")

```
###############################################################################################
# exclusion codes
###############################################################################################

```{r get unique preg canc codes}

icd10_cancer <- unique(c(icd10_canc_oct$DX_CODE_2, icd10_canc_nov$CONCEPT_CODE))
  
icd9_cancer <- unique(icd9_canc_oct$DX_CODE_2)

icd10_pregnancy <- unique(c(icd10_preg_nov$CONCEPT_CODE, icd10_preg_oct$DX_CODE_2))
  
icd9_pregnancy <- unique(icd9_preg_oct$DX_CODE_2)
  
```

```{r make icd9 and icd10 code lists}

icd9_pc <- c(icd9_cancer, icd9_pregnancy)

icd10_pc <- c(icd10_cancer, icd10_pregnancy)

```

```{r}

icd9_phe <- phecode_map[phecode_map$ICD %in% icd9_pc & phecode_map$Flag == 9,]$Phecode

icd10_phe <- phecode_map[phecode_map$ICD %in% icd10_pc & phecode_map$Flag == 10,]$Phecode

phe_pc <- unique(c(icd9_phe, icd10_phe))

```

```{r}

phecodes_icd10 <- phecodes[phecodes$icd == 10 & phecodes$code %in% icd10_pc,]

phecodes_icd9 <- phecodes[phecodes$icd == 9 & phecodes$code %in% icd9_pc,]

phecodes_pc <- rbind(phecodes_icd10, phecodes_icd9)


rm(phecodes_icd10, phecodes_icd9)

```


Phecodes had man non-specific diagnosis terms
```{r}

phecodes_pc <- phecodes[phecodes$Phecode %in% phe_pc,]

```

Persons with potential health hazards related to socioeconomic, psychosocial, and other circumstances

```{r}

phecodes_pc <- phecodes_pc[phecodes_pc$Phecode != "1010.7", ]

```

###############################################################################################
# head_circumference
###############################################################################################

```{r}

hc_df <- encounter_df[!is.na(encounter_df$head_circumference),]

```

```{r}

hc_df_birth <- hc_df[hc_df$effective_age == 0 , ]

```
3 indv with multiple birth measurements
   1    2 
1907    3 
they are about the same; +/- 0.5cm

```{r}

hc_df_birth_largest <- hc_df_birth %>%
  group_by(pat_id) %>%
  slice_max(order_by = head_circumference, n = 1, with_ties = FALSE) %>%
  ungroup()


```

```{r}

birth_headc_hist <- hc_df_birth_largest %>%
  ggplot( aes(x=head_circumference )) +
    geom_histogram( binwidth=1, fill="blue3", color="#e9ecef", alpha=0.9) +
    ggtitle("Head Circumference at Birth (cm)") +
    theme(
      plot.title = element_text(size=15)
    )


```

```{r}

ggsave("birth_headc_hist.pdf", plot = birth_headc_hist, width = 6, height = 4, units = "in")

```

###############################################################################################
# bmi
###############################################################################################
weight(kg)/height(m)^2

```{r}

encounter_df$bmi <- encounter_df$weight_kg/(encounter_df$height_cm/100)^2

```

```{r}

bmi_mean <- mean(encounter_df$bmi, na.rm = TRUE)
bmi_sd   <- sd(encounter_df$bmi, na.rm = TRUE)


```

```{r}

encounter_df <- encounter_df %>%
  mutate(bmi_pruned = ifelse(
    bmi >= bmi_mean - 6 * bmi_sd & bmi <= bmi_mean + 6 * bmi_sd,
    bmi,
    NA
  ))


```

```{r}

total_bmi_plot <- ggplot(encounter_df, aes(x = effective_age, y = bmi_pruned, group = pat_id, color = factor(pat_id))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  geom_vline(xintercept = 6574.5, linetype = "dashed", color = "black", linewidth = 1) +
  labs(x = "Age", y = "BMI", color = "Individual") +
  theme_minimal() +
  theme(legend.position = "none")


```

```{r}

mean_max_age_bmi <- encounter_df %>%
  # keep only rows with non-missing BMI
  filter(!is.na(bmi_pruned)) %>%
  # group by individual
  group_by(pat_id) %>%
  # get each person's maximum effective_age
  summarize(max_age = max(effective_age, na.rm = TRUE)) %>%
  # compute the mean of these maximum ages
  summarize(mean_max_age_bmi = mean(max_age, na.rm = TRUE)) %>%
  pull(mean_max_age_bmi)

```

```{r}

ggsave("raw_bmi_plot.pdf", plot = total_bmi_plot, width = 8, height = 6, units = "in")

```

```{r}


encounter_df <- encounter_df %>%
  mutate(bmiz_pruned = ifelse(
    bmiAgeZ >= - 6  & bmiAgeZ <= 6,
    bmiAgeZ,
    NA
  ))


```

```{r}

bmiz_plot <- ggplot(encounter_df, aes(x = effective_age, y = bmiz_pruned, group = pat_id, color = factor(pat_id))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  geom_vline(xintercept = 6574.5, linetype = "dashed", color = "black", linewidth = 1) +
  labs(x = "Age", y = "BMI-Age-Z", color = "Individual") +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r}

ggsave("bmiz_plot.pdf", plot = bmiz_plot, width = 8, height = 6, units = "in")

```

```{r}

headc_mean <- mean(encounter_df$head_circumference, na.rm = TRUE)
headc_sd   <- sd(encounter_df$head_circumference, na.rm = TRUE)


```

```{r}

encounter_df <- encounter_df %>%
  mutate(head_circumference_pruned = ifelse(
    head_circumference >= headc_mean - 6 * headc_sd & head_circumference <= headc_mean + 6 * headc_sd,
    head_circumference,
    NA
  ))


```

```{r}

headc_plot <- ggplot(encounter_df, aes(x = effective_age, y =head_circumference_pruned , group = pat_id, color = factor(pat_id))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  geom_vline(xintercept = 6574.5, linetype = "dashed", color = "black", linewidth = 1) +
  labs(x = "Age (Days)", y = "Head Circumference (cm)", color = "Individual") +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r}

ggsave("head_circ_total.pdf", plot = headc_plot, width = 8, height = 6, units = "in")

```


```{r}

encounter_df <- encounter_df %>%
  mutate(hcz_pruned = ifelse(
    hcAgeZ  >= -6 & hcAgeZ <=  6,
    hcAgeZ ,
    NA
  ))


```

```{r}

headcz_plot <- ggplot(encounter_df, aes(x = effective_age, y =hcz_pruned , group = pat_id, color = factor(pat_id))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  xlim(-1, (5*365.25)) +
 # geom_vline(xintercept = 6574.5, linetype = "dashed", color = "black", linewidth = 1) +
  labs(x = "Age (Days)", y = "Head Circumference Z-score", color = "Individual") +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r}

bmiz__prune_spag <- ggplot(encounter_df, aes(x = effective_age, y = bmiz_pruned, group = pat_id)) +
  geom_line(alpha = 0.1, color = "steelblue") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Individual BMI trajectories")

```

```{r}

bmiz_freq <- table(encounter_df$pat_id[!is.na(encounter_df$bmiz_pruned)])


```

```{r}

bmiz__prune_spag_NG0MNYETRS86L <- ggplot(encounter_df[encounter_df$pat_id == "NG0MNYETRS86L",], aes(x = effective_age, y = bmiz_pruned, group = pat_id)) +
  geom_line(alpha = 1, color = "steelblue") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Individual BMI trajectories (sample)")

```

```{r}

bmiz_plot_NG0MNYETRS86L <- ggplot(encounter_df[encounter_df$pat_id == "NG0MNYETRS86L",], aes(x = effective_age, y = bmiz_pruned, group = pat_id, color = factor(pat_id))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  geom_vline(xintercept = 6574.5, linetype = "dashed", color = "black", linewidth = 1) +
  labs(x = "Age", y = "BMI-Age-Z", color = "Individual") +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r}

ggsave("indv_1_bmiz_age.pdf", plot = bmiz_plot_NG0MNYETRS86L, width = 8, height = 6, units = "in")

```

```{r}

bmi_plot_NG0MNYETRS86L <- ggplot(encounter_df[encounter_df$pat_id == "NG0MNYETRS86L",], aes(x = effective_age, y = bmi, group = pat_id, color = factor(pat_id))) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  geom_vline(xintercept = 6574.5, linetype = "dashed", color = "black", linewidth = 1) +
  labs(x = "Age", y = "BMI", color = "Individual") +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r}

ggsave("indv_1_bmi_age.pdf", plot = bmi_plot_NG0MNYETRS86L, width = 8, height = 6, units = "in")

```

```{r}

# Find the largest effective_age per pat_id
max_age_df <- encounter_df %>%
  group_by(pat_id) %>%
  summarize(max_effective_age = max(effective_age, na.rm = TRUE))

# View result
head(max_age_df)

```